Red Wine quality assessment
Data analytics
03/15/2018
Submitted by Vikas Warudkar
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE}
require(ggplot2)
require(MASS)
require(gridExtra)
require(dplyr)
require(memisc)
require(GGally)
require(reshape2)
require(kernlab)
update.packages("ggplot2")
library(plyr)

```

Read csv file and explore statistics
```{r echo=FALSE,message=FALSE,warning=FALSE}
Wine <- read.csv('wineQualityReds.csv')
str(Wine) 
summary(Wine)
Wine$quality <- as.numeric(Wine$quality)

```

Creates tabular results of categorical variables

```{r,message=FALSE,warning=FALSE}
table(Wine$quality)

```


# Univariate Plots Section

```{r echo=FALSE,message=FALSE,warning=FALSE}
grid.arrange(qplot(Wine$fixed.acidity),
             qplot(Wine$volatile.acidity),
             qplot(Wine$citric.acid),
             qplot(Wine$residual.sugar),
             qplot(Wine$chlorides),
             qplot(Wine$free.sulfur.dioxide),
             qplot(Wine$total.sulfur.dioxide),
             qplot(Wine$density),
             qplot(Wine$pH),
             qplot(Wine$sulphates),
             qplot(Wine$alcohol),
             qplot(Wine$quality),
             ncol = 4)

```

# Univariate Analysis

1. Wine Quality forms a normal distribution.

2. Density and pH are normally distributed with a few outliers.

Create new variable for better exploration
```{r,message=FALSE,warning=FALSE}
Wine$rating <- ifelse(Wine$quality < 5, 'bad', ifelse(
  Wine$quality < 7, 'average', 'good'))
Wine$rating <- ordered(Wine$rating,
                     levels = c('bad', 'average', 'good'))
summary(Wine$rating)
```


Create Histogram of log function of the variables for further analysis
```{r,message=FALSE,warning=FALSE}
ggplot(Wine,aes(x=fixed.acidity))+geom_histogram(fill='red')+scale_x_log10(breaks=4:15)+
  xlab('Fixed Acidity')+ylab('Count')+ggtitle('Histogram of Fixed Acidity Values')
require(plotly)
help(package='plotly')
ggplot()
plot_ly(data=Wine,x=~citric.acid,type='histogram')

ggplot(Wine) + geom_histogram(aes(x=volatile.acidity),fill='blue')+
  scale_x_log10(breaks=seq(0.1,1,0.1))

ggplot(Wine) + 
  geom_histogram(aes(x=citric.acid),fill='green') +
  scale_x_log10()
```

Citric acid was one feature that was found to be not 
normally distributed on a logarithmic scale. 


Create a combined variable, 
TAC.acidity, containing the sum of tartaric, acetic, and citric acid.

```{r,message=FALSE,warning=FALSE}
Wine$TAC.acidity <- Wine$fixed.acidity + Wine$volatile.acidity + 
Wine$citric.acid

qplot(Wine$TAC.acidity,main = 'Histogram of TAC Acidity (fixed+volatile+Citric)')

```

## Boxplots are better suited in visualizing the outliers.

```{r,message=FALSE,warning=FALSE}
get_simple_boxplot <- function(column, ylab) {
  return(qplot(data = Wine, x = 'simple',
               y = column, geom = 'boxplot',
               xlab = '',
               ylab = ylab))
}

grid.arrange(get_simple_boxplot(Wine$fixed.acidity, 'fixed acidity'),
             get_simple_boxplot(Wine$volatile.acidity, 'volatile acidity'),
             get_simple_boxplot(Wine$citric.acid, 'citric acid'),
             get_simple_boxplot(Wine$TAC.acidity, 'TAC acidity'),
             get_simple_boxplot(Wine$residual.sugar, 'residual sugar'),
             get_simple_boxplot(Wine$chlorides, 'chlorides'),
             get_simple_boxplot(Wine$free.sulfur.dioxide, 'free sulf. dioxide'),
             get_simple_boxplot(Wine$total.sulfur.dioxide, 'total sulf. dioxide'),
             get_simple_boxplot(Wine$density, 'density'),
             get_simple_boxplot(Wine$pH, 'pH'),
             get_simple_boxplot(Wine$sulphates, 'sulphates'),
             get_simple_boxplot(Wine$alcohol, 'alcohol'),
             ncol = 4)

plot_ly(Wine,y=~alcohol,type='box')

```

# Bivariate Plots Section

```{r echo=FALSE,message=FALSE,warning=FALSE}

get_bivariate_boxplot <- function(x, y, ylab) {
  return(qplot(data = Wine, x = x, y = y, geom = 'boxplot', ylab = ylab))
}

grid.arrange(get_bivariate_boxplot(Wine$quality, Wine$fixed.acidity,
                                   'fixed acidity'),
             get_bivariate_boxplot(Wine$quality, Wine$volatile.acidity,
                                   'volatile acidity'),
             get_bivariate_boxplot(Wine$quality, Wine$citric.acid,
                                   'citric acid'),
             get_bivariate_boxplot(Wine$quality, Wine$TAC.acidity,
                                   'TAC acidity'),
             get_bivariate_boxplot(Wine$quality, log10(Wine$residual.sugar),
                                   'residual sugar'),
             get_bivariate_boxplot(Wine$quality, log10(Wine$chlorides),
                                   'chlorides'),
             get_bivariate_boxplot(Wine$quality, Wine$free.sulfur.dioxide,
                                   'free sulf. dioxide'),
             get_bivariate_boxplot(Wine$quality, Wine$total.sulfur.dioxide,
                                   'total sulf. dioxide'),
             get_bivariate_boxplot(Wine$quality, Wine$density,
                                   'density'),
             get_bivariate_boxplot(Wine$quality, Wine$pH,
                                   'pH'),
             get_bivariate_boxplot(Wine$quality, log10(Wine$sulphates),
                                   'sulphates'),
             get_bivariate_boxplot(Wine$quality, Wine$alcohol,
                                   'alcohol'),
             ncol = 4)


```

Correlation for each of these
variables against quality:

```{r,message=FALSE,warning=FALSE}
simple_cor_test <- function(x, y) {
  return(cor.test(x, as.numeric(y))$estimate)
}

correlations <- c(
  simple_cor_test(Wine$fixed.acidity, Wine$quality),
  simple_cor_test(Wine$volatile.acidity, Wine$quality),
  simple_cor_test(Wine$citric.acid, Wine$quality),
  simple_cor_test(Wine$TAC.acidity, Wine$quality),
  simple_cor_test(log10(Wine$residual.sugar), Wine$quality),
  simple_cor_test(log10(Wine$chlorides), Wine$quality),
  simple_cor_test(Wine$free.sulfur.dioxide, Wine$quality),
  simple_cor_test(Wine$total.sulfur.dioxide, Wine$quality),
  simple_cor_test(Wine$density, Wine$quality),
  simple_cor_test(Wine$pH, Wine$quality),
  simple_cor_test(log10(Wine$sulphates), Wine$quality),
  simple_cor_test(Wine$alcohol, Wine$quality))

correlations

names(correlations) <- c('fixed.acidity', 'volatile.acidity', 'citric.acid',
                         'TAC.acidity', 'log10.residual.sugar',
                         'log10.chlordies', 'free.sulfur.dioxide',
                         'total.sulfur.dioxide', 'density', 'pH',
                         'log10.sulphates', 'alcohol')
correlations
```

Top 4:
alcohol
sulphates (log10)
volatile acidity
citric acid

Examining the acidity variables:

```{r,message=FALSE,warning=FALSE}
ggplot(data = Wine, aes(x = fixed.acidity, y = citric.acid)) +
  geom_point(alpha=0.3)

cor.test(Wine$fixed.acidity, Wine$citric.acid)

ggplot(data = Wine, aes(x = volatile.acidity, y = citric.acid)) +
  geom_point(alpha=0.3)

cor.test(Wine$volatile.acidity, Wine$citric.acid)

ggplot(data = Wine, aes(x = log10(TAC.acidity), y = pH)) +
  geom_point(alpha=0.3)

cor.test(log10(Wine$TAC.acidity), Wine$pH)

```

Base 10 logarithm TAC.acidity correlated very well with pH.

Building a predictive linear model, 
to predict pH based off of TAC.acidity and
capture the % difference as a new variable.

```{r,message=FALSE,warning=FALSE}

m <- lm(I(pH) ~ I(log10(TAC.acidity)), data = Wine)

Wine$pH.predictions <- predict(m, Wine)

# (observed - expected) / expected
Wine$pH.error <- (Wine$pH.predictions - Wine$pH)/Wine$pH

```

To check its accuracy.
The RMS Error.

```{r,message=FALSE,warning=FALSE}
rmse <- function(error)
{
    sqrt(mean(error^2))
}

rmse(m$residuals)
#Now, we train a Support Vector Machine.

require(e1071)

SVM <- svm(I(pH) ~ I(log10(TAC.acidity)), data = Wine)

Wine$pH.Predict.SVM <- predict(SVM,Wine)

Wine$pH.error.SVM <- (Wine$pH.Predict.SVM - Wine$pH)/Wine$pH

rmse(SVM$residuals)


```

SVM functions slightly better than a LM.

### Plot 1: Effect of Alcohol on Wine Quality
```{r echo=FALSE,message=FALSE,warning=FALSE}
ggplot(data = Wine, aes(x = quality, y = alcohol,
                      fill = rating)) +
  geom_boxplot(outlier.color = 'red') +
  ggtitle('Alcohol Levels in Different Wine Qualities') +
  xlab('Quality') +
  ylab('Alcohol (% volume)')

```

### Description 1
These boxplots demonstrate the effect of alcohol content on wine quality.
Generally, higher alcohol content correlated with higher wine quality.
However, as the outliers and intervals show, alchol content alone did not
produce a higher quality.
